\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-12 paper.qmd
Added images of handwritten digits

```{r}
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)

@@

$$
 y_{max} = \text{ceiling}(\frac{ar}{hr}) * hr
$$ {#eq-equation3}

@@

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for PBMC3k dataset (a) colored by manualy identified clusters, (b) colored by `detour` results"
#| label: fig-nldervisPBMCUMAP3Res
#| out-width: 100%

@@

@@

@@

plot_list2_mnist <- pacmap_minst_scaled |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

@@

## To draw with AIC
aic_plot_mnist <- ggplot(mse_df_mnist, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 80, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_mnist <- ggplot(mse_df_mnist, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 80, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from PaCMAP applied to training MNIST dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 80, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltMNIST
##| out-width: 100%
#| fig-pos: H

@@

## Decide by looking at MSE plot
num_bins_x_mnist <- 8
num_bins_y_mnist <- 10
hex_size_mnist <- 0.1

@@

trimesh_removed_mnist_pacmap <- trimesh_removed_mnist_pacmap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 2, "error 0-2",
                                if_else(total <= 4, "error 2-4",
                                        if_else(total <= 6, "error 4-6",
                                                if_else(total <= 8, "error 6-8",
                                                        if_else(total <= 12, "error 8-10",
                                                                if_else(total <= 14, "error 10-12", "error greter than 12"))))))))

plot_list2_mnist_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-2", "error 2-4", "error 4-6", "error 6-8", "error 8-10", "error 10-12", "error greter than 12"))) |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d"))

@@

>>>>

```{r}
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)
library(tidyr)

@@

$$
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
$$ {#eq-equation3}

@@

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for PBMC3k dataset (a) colored manually in 2D, (b) colored in high-D"
#| label: fig-nldervisPBMCUMAP3Res
#| out-width: 100%

@@

- Selected 10 PCs based on scree plot

- 7877 images of digit 1

@@

#### Model

- Total number of bins used: 180 (num_x = 12, num_y = 15, hex_size = 0.06)

- Learned from the model: There is a non-linear shape in 2D, but if look at the model in high-D, the non-linear continuity is there but there is a twist in the structure which didn't capture by PaCMAP.

- Write about looking at images of digit 1 as well..how in 2D space images are distributed, which images occur high-errors in high-D

@@

plot_list2_mnist <- pacmap_minst_scaled |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3) 
```

@@

## To draw with AIC
aic_plot_mnist <- ggplot(mse_df_mnist, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 180, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_mnist <- ggplot(mse_df_mnist, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 180, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from PaCMAP applied to training MNIST dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 180, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltMNIST
##| out-width: 100%
#| fig-pos: H

@@

## Decide by looking at MSE plot
num_bins_x_mnist <- 12
num_bins_y_mnist <- 15
hex_size_mnist <- 0.06

@@

trimesh_removed_mnist_pacmap <- trimesh_removed_mnist_pacmap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 2, "error 0-2",
                                if_else(total <= 4, "error 2-4",
                                        if_else(total <= 6, "error 4-6",
                                                if_else(total <= 10, "error 6-10",
                                                        if_else(total <= 12, "error 10-12",
                                                                if_else(total <= 14, "error 12-14", "error greter than 14"))))))))

plot_list2_mnist_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-2", "error 2-4", "error 4-6", "error 6-10", "error 10-12", "error 12-14", "error greter than 14"))) |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) +
  annotate(geom = 'text', label = 'c', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

<!--images with large error in high-D-->

img_error <- error_df |> 
  dplyr::filter(total >= 10) |> 
  dplyr::pull(ID)

mnist_data <- read_rds("data/mnist/mnist_digit_1.rds")
pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% img_error)

theme_set(theme_light())
imge_error_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank())
```

```{r}
#| echo: false
#| fig-cap: "Images of handwritten digit 1 which occur large model error."
#| label: fig-sampleimg
#| out-width: 100%
#| fig-pos: H

imge_error_sample
```

### 2024-03-12 paper.tex
Added images of handwritten digits

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 y_{max} = \text{ceiling}(\frac{ar}{hr}) * hr
}\label{eq-equation3}\end{equation}

@@

\caption{\label{fig-nldervisPBMCUMAP3Res}2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for PBMC3k
dataset (a) colored by manualy identified clusters, (b) colored by
\texttt{detour} results}

@@

\begin{itemize}
\item
  Scale the pixel values to the range {[}0,1{]}
\item
  Reshape to have a 2-dimensional array
\item
  reshaping pixel data is a fundamental preprocessing step in image
  analysis tasks, ensuring that the data is in a suitable format for
  further analysis, modeling, and interpretation. It helps preserve
  important spatial information, ensures compatibility with machine
  learning models, and facilitates feature extraction and visualization.
\item
  Reshaping the pixel data allows us to preserve this spatial
  information, ensuring that the relationships between neighboring
  pixels are maintained.
\item
  Many machine learning algorithms, including neural networks, expect
  input data to be in a specific format. For image data, this typically
  means a two-dimensional array where each row represents a single
  sample (image), and each column represents a feature (pixel value).
  Reshaping ensures that all samples have consistent dimensions, making
  them suitable for feeding into these algorithms.
\item
  Reshaping the pixel data allows for easier extraction of relevant
  features from the images. Once reshaped, individual pixels can be
  treated as features, making it easier to apply techniques like
  dimensionality reduction (e.g., PCA) or feature engineering to extract
  meaningful information from the images.
\item
  PaCMAP parameters: n\_components=2, n\_neighbors=10, init=``random'',
  MN\_ratio=0.9, FP\_ratio=2.0
\end{itemize}

@@

\caption{\label{fig-diagnosticpltMNIST}Goodness of fit statistics from
PaCMAP applied to training MNIST dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
80, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\begin{verbatim}
$mse
[1] 3.017065

$aic
[1] 87724.49
\end{verbatim}

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
}\label{eq-equation3}\end{equation}

@@

\caption{\label{fig-nldervisPBMCUMAP3Res}2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for PBMC3k
dataset (a) colored manually in 2D, (b) colored in high-D}

@@

\begin{itemize}
\item
  Scale the pixel values to the range {[}0,1{]}
\item
  Reshape to have a 2-dimensional array
\item
  reshaping pixel data is a fundamental preprocessing step in image
  analysis tasks, ensuring that the data is in a suitable format for
  further analysis, modeling, and interpretation. It helps preserve
  important spatial information, ensures compatibility with machine
  learning models, and facilitates feature extraction and visualization.
\item
  Reshaping the pixel data allows us to preserve this spatial
  information, ensuring that the relationships between neighboring
  pixels are maintained.
\item
  Many machine learning algorithms, including neural networks, expect
  input data to be in a specific format. For image data, this typically
  means a two-dimensional array where each row represents a single
  sample (image), and each column represents a feature (pixel value).
  Reshaping ensures that all samples have consistent dimensions, making
  them suitable for feeding into these algorithms.
\item
  Reshaping the pixel data allows for easier extraction of relevant
  features from the images. Once reshaped, individual pixels can be
  treated as features, making it easier to apply techniques like
  dimensionality reduction (e.g., PCA) or feature engineering to extract
  meaningful information from the images.
\item
  Selected 10 PCs based on scree plot
\item
  7877 images of digit 1
\item
  PaCMAP parameters: n\_components=2, n\_neighbors=10, init=``random'',
  MN\_ratio=0.9, FP\_ratio=2.0
\end{itemize}

\hypertarget{model}{%
\subsubsection{Model}\label{model}}

\begin{itemize}
\item
  Total number of bins used: 180 (num\_x = 12, num\_y = 15, hex\_size =
  0.06)
\item
  Learned from the model: There is a non-linear shape in 2D, but if look
  at the model in high-D, the non-linear continuity is there but there
  is a twist in the structure which didn't capture by PaCMAP.
\item
  Write about looking at images of digit 1 as well..how in 2D space
  images are distributed, which images occur high-errors in high-D
\end{itemize}

@@

\caption{\label{fig-diagnosticpltMNIST}Goodness of fit statistics from
PaCMAP applied to training MNIST dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
180, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\begin{verbatim}
$mse
[1] 2.276023

$aic
[1] 66382.78
\end{verbatim}

@@

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sampleimg-1.pdf}

\caption{\label{fig-sampleimg}Images of handwritten digit 1 which occur
large model error.}
